Hydras and Cie (Draft) \\ Variations on Kirby \& Paris' hydra battles and ordinal numbers in Coq

### 2021-09-19 doc/JFLA2022/paper.tex
minor corrections

\begin{abstract}
  We present the current state of a \textit{Coq} library on ordinal numbers.

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description  can be found in Andrej Bauer's blogpost~\cite{bauer2008} as well as a simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\vref{fig:iota-example}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $i$ of times. The number $i$ of copies and the copied sub-hydra $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
\end{itemize}

@@

   \begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.3]
\node (foot) at (10,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (10,2) {$\bullet$};
\node (N22) at (7,2) {$\bullet$};
\node (N3) at (14,2) {$\bullet$};
\node (N4) at (18,2) {$\Smiley[2][green]$};
\node (N5) at (0,4) {$\bullet$};
\node (N6) at (2,5) {$\Smiley[2][green]$};
\node (N7) at (4,6) {$\Smiley[2][green]$};
\node (N88) at (7,4) {$\bullet$};
\node (N8) at (10,4) {$\bullet$};
\node (N9) at (14,6) {$\Smiley[2][green]$};
\node (N10) at (0,8) {$\Smiley[2][green]$};
\node (N11) at (10,7) {$\Smiley[2][green]$};
\node (N111) at (7,7) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N22);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (N1) to  (N5);
\draw (N1) to   [bend left=10] (N6);
\draw (N1) to   [bend right=20] (N7);
\draw (N2) to  (N8);
\draw (N22) to  (N88);
\draw (N8) to  (N11);
\draw (N88) to  (N111);
\draw (N3) to  (N9);
\draw (N5) to  (N10);
\end{tikzpicture}
\caption{The hydra associated with $\omega^{\omega+2}+\omega^\omega \times 2 + \omega + 1$ \label{fig:iota-example}}

@@

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-Community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementaion in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic) definitions of
\texttt{T1} and \texttt{T2}, originated in~\cite{CantorContrib}.
The types --- defined in different modules --- cannot be considered as equal. Even more, some operations like addition, multiplication, etc. have now different definitions.

It would be interesting to build a ``gate'' giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}.
For instance, the associativity of multiplication of ordinals below $\epsilon_0$, already proven in \textit{Gaia} was tranferred ``for free'' to the data structures of \textit{hydra-battles}, thanks to the isomorphism between both implementations.

Another direction would be to write in \textit{Coq} a formal proof of the statement of the second theorem of Kirby and Paris, using the formalization of Peano Arithmetic by O'Connor~\cite{Goedel}.

>>>>

\begin{abstract}
  \textbf{To do!} We present the current state of a \textit{Coq} library on ordinal numbers.

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description  can be found in Andrej Bauer's blogpost~\cite{bauer2008} as well as a simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\vref{fig:iota-example}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $i$ of times. The number $i$ of copies and the copied sub-hydra $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
\end{itemize}
   \begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.3]
\node (foot) at (10,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (10,2) {$\bullet$};
\node (N22) at (7,2) {$\bullet$};
\node (N3) at (14,2) {$\bullet$};
\node (N4) at (18,2) {$\Smiley[2][green]$};
\node (N5) at (0,4) {$\bullet$};
\node (N6) at (2,5) {$\Smiley[2][green]$};
\node (N7) at (4,6) {$\Smiley[2][green]$};
\node (N88) at (7,4) {$\bullet$};
\node (N8) at (10,4) {$\bullet$};
\node (N9) at (14,6) {$\Smiley[2][green]$};
\node (N10) at (0,8) {$\Smiley[2][green]$};
\node (N11) at (10,7) {$\Smiley[2][green]$};
\node (N111) at (7,7) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N22);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (N1) to  (N5);
\draw (N1) to   [bend left=10] (N6);
\draw (N1) to   [bend right=20] (N7);
\draw (N2) to  (N8);
\draw (N22) to  (N88);
\draw (N8) to  (N11);
\draw (N88) to  (N111);
\draw (N3) to  (N9);
\draw (N5) to  (N10);
\end{tikzpicture}
\caption{The hydra associated with $\omega^{\omega+2}+\omega^\omega \times 2 + \omega + 1$ \label{fig:iota-example}}

@@

@@

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-Community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic but not identical) definitions of types \texttt{T1} and \texttt{T2}, both originated in~\cite{CantorContrib}.
Our plan is to build a  ``gate'' giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems from one of the two libraries to the other one.

Another direction would be to write in \textit{Coq} a formal proof of the original statement of the second theorem of Kirby and Paris. For this purpose, we plan to import  the formalization of Peano Arithmetic by O'Connor~\cite{Goedel}.
